Research Experience

### 2010-06-25 doc/thesis/thesis.tex
"finished" kalman filter section

\documentclass{article}
\usepackage[left=3cm,top=2cm,right=3cm,nohead,nofoot]{geometry}

\usepackage{times}
\usepackage[siunitx]{circuitikz}
%\usetikzlibrary{trees}
%\usetikzlibrary{shapes,snakes}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{cancel}
\usepackage{hyperref}

@@

\begin{document}
\section{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
of neural activity. Despite its rather limited temporal resolution, FMRI is still
the best way of measuring neural activity for the majority of the brain.
Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis straight forward.
Because of these benefits, FMRI continues to be crucial to the study of human 
cognition. Despite its prevalence, there have been relatively few developments
in the actual analysis of FMRI images. A steady stream of studies have built
on the original BOLD signal derivation first described in \cite{Ogawa}, 
from the Baloon model first proposed by \cite{Buxton1998}
all the way to full fully autonomous system of equations \cite{Riera2004}. And while
there have been numerous forks in the model, enough in fact to make an entire paper
studying the differences,\cite{Deneux2006}, it is widely known that all these
models have quantitatively less bias error than General Linear Model which is
typically employed today. Then again, depending on the model there may be between
seven \cite{Riera2004} and 50 \cite{Behzadi2005} parameters per voxel to
be optimized. Clearly there is a significant risk of error due to variance
with so many degrees of freedom, not to mention a significantly increased
computation cost. In this thesis I demonstrate the use of a 
particle filter as a means of addressing these problems.

@@

\subsection{FMRI}
Magnetic Resonance Imaging, MRI is a method of building 3D images
non-invasively, based on the difference between nuclear spin
relaxation times in various molecules. Initially the entity being
imaged is brought into a large magnetic field which aligns the spins
of molecules in the same direction; radio frequency (RF) signals may
then be used to excite nuclear spin away from the steady alignment. 
As the nuclei precess back to their original orientation, they resonate
at the same RF frequency of their original excitation. Conveniently, the
excitation of nuclear spins return their original state at different
rates, called the T1 relaxation time, depending on the properties of 
the material excited. Additionally, the
coherence of the spins also decay differently (and quite a bit faster
than T1) based on the properties of the region that has been excited.
This gives two primary methods of contrasting substances,
which is the basis of T1 and T2 weighted images. Additionally, there
dephasing occurs at two different rates, the T2 relaxation time,
which is impossible to recover from, and T2$^*$ relaxation, which is
much faster, but possible to recover from with an inversion pulse.
Oftentimes T1 relaxation times can be on the order of seconds if 
a significant excitation pulse is applied. 
In order to rapidly acquire entire brain images, as is done in Functional 
MRI, a single large excitation pulse is applied to the entire brain,
and the entire volume is acquired in a single T1 relaxation period. 
Because the entire k-space (spatial-frequency) volume is acquired 
from a single excitation, the signal to noise ration is very low
in this type  of imaging (Echo Planar Imaging).

@@

\subsection{BOLD Physiology}
It is well known that the two types of hemoglobin act as a contrast agents in 
EPI imaging
\cite{Buxton1998}, \cite{WEISSKOFF1994}, \cite{Ogawa}, however the connection
between Deoxyhemoglobin/Oxygenated Hemoglobin and neural activity is non-trivial. 
Intuitively, increased 
metabolism will increase Deoxyhemoglobin, however blood vessels are quick
to compensate by increasing local blood flow. Increased inflow will of course
preceed increased outflow, and increased inflow is accomplished by loosening 
capilary beds. Both of these factors drive increased storage capacity.
Since the local MR signal depends on the ratio of Deoxyhemoglobin to Oxygenated
Hemoglobin, increased volume of blood can certainly effect this ratio if 
metabolism doesn't exactly match the increased inflow of oxygenated blood.
This was the impetus
for the ground breaking balloon model (\cite{Buxton1998}) and windkessel
model (\cite{Mandeville1999}). These models derive from first principals
the increased deoxyhemoglobin ratio and volume of capillaries based on a given flow.
These were the first two attempts to quantitatively account for the shape of the 
BOLD signal as a consequence of the lag between the cerebral blood volume (CBV) 
and the cerebral blood flow (CBF). In fact \cite{Buxton1998} went to far as
to show that a simple, well chosen blood flow waveform coupled with a square 
wave cerebral metabolic rate of oxygen (CMRO2) curve, in the context of a balloon 
model, could fully account for the BOLD signal.

Although \cite{Buxton1998} showed that a well chosen flow waveform could 
explain much of the BOLD signal, there was still a matter of proposing a
realistic waveform for the CBF and for the CMRO2. \cite{friston2000} gave
a reasonable and simple
expressoin for CBF input,$f$, based on a flow inducing signal, $s$, 
\begin{eqnarray}
\dot{s} &=& \epsilon u(t) - \frac{s}{\tau_s} - \frac{f - 1}{\tau_f} \\
\dot{f} &=& s
\end{eqnarray}
where $\epsilon$ is a neuronal efficiency term, $u(t)$ is a stimulus, and $\tau_f$, $\tau_s$
are both time constants. In \cite{Buxton2004} the final piece of the simple balloon
model was put into place, by describing the CMRO2 as a constant multiple of
the CBF (the inflow of blood). This completed the basic balloon model, and 
was well summarized in \cite{Riera2004}. 
\begin{eqnarray}
\dot{v} &=& \frac{1}{\tau_0}(f - v^\alpha)\\
\dot{q} &=& \frac{1}{\tau_0}(\frac{f(1-(1-E_0)^f)}{E_0} - \frac{q}{v^{1-1/\alpha}})
\end{eqnarray}
where $v$ is normalized cerebral blood volume (CBV), and $q$ is the normalized
local deoxyhemoglobin/oxygenated hemoglobin ratio, $E_0$ is the resting metabolic
rate and $\alpha$ is Grubb's parameter controling the balloon model. 
\cite{Obata2004} refined the readout equation of the BOLD signal based on the
deoxyhemoglobin content (q) and local blood volume (v), resulting in the
final BOLD equation:
\begin{eqnarray}
y   &=& V_0((k_1 + k_2)(1-q) - (k_2 + k_3)(1-v))\\
k_1 &=& 4.3 \times \nu_0 \times E_0 \times TE = 2.8\\
K_2 &=& \epsilon_0 \times r_0 \times E_0 \times TE = .57\\
k_3 &=& \epsilon_0 - 1 = .43
\end{eqnarray}
Where $\nu_0 = 40.3 s^{-1}$  is the frequency offset in Hz for fully
deoxygenated blood (at 1.5T), $r_0 = 25 s^{-1}$  is the slope relating
change in relaxation rate with change in blood oxygenation, and
$\epsilon_0 = 1.43$ is the 
ratio of signal MR from intravascular to extravascular at rest. Although,
obviously these constants change with experiment ($TE$, $\nu_0$, $r_0$),
patient, and brain 
region ($E_0$, $r_0$), often the estimated values taken from \cite{Obata2004} are used
as constants ($k_1 + k_2 = 3.4$, and $k_2+k_3 = 1$) in 1.5 Tesla studies..
While this model is in a sense complete, it is far from perfect. The major
problem often brought up with this version of the BOLD model is that it
does not represent the so called "post-stimulus undershoot" well.
The post-stimulus undershoot is the name for a prolonged sub-normal
BOLD response for a period of 10 to 60 seconds after stimulus has
ceased (\cite{Chen2009}, \cite{Mandeville1999a}).

@@

Even more advanced versions of the Balloon model exist. In fact \cite{Buxton2004}
introduced several additional state variables, including the CMRO2, the O2 extraction
fraction, which is closely related to CMR02, and the neural response, which 
causes the stimulus to decay toward some steady state value. The 
neural response is intended to emulate neural habituation, wherein neurons
become less sensitive to a prolonged stimulus. While these advanced may be more
capable of capturing a more exact version of the BOLD signal, the difference
between the model will often times be below the noise floor. In essence this 
is a classic bias-variance
dilemma: at some point increased model flexibility, and thus variance, is not
worth the decrease in model bias. For now it remains to be seen where this
line may be drawn in the BOLD signal, although \cite{deneux2006} does not
show a significant improvement from the additional parameters added in 
\cite{Buxton2004}.

\subsection{Previous Studies of Parameters}
There have been quite a few efforts to quantify the parameters of the
various BOLD models.  Although \cite{Buxton1998} and \cite{Friston2000}
both proposed physiologically reasonable values for the model parameters, 
\cite{Friston2002} was the first paper to calculate the parameters based 
on actual FMRI data. In that paper, Friston et. al. used a variation of
Expectation Maximization to find a normal distribution for the parameters:
\begin{eqnarray}
\epsilon &=& N(.54 , .1 ^2 )  \nonumber \\
\tau_s & =&  N(1.54, .25^2)   \nonumber \\
\tau_f & =&  N(2.46, .25^2)   \nonumber \\
\tau_0 & =&  N(.98 , .25^2 )   \nonumber \\
\alpha & =&  N(.33 , .45^2 )   \nonumber \\
E_0   & =&  N(.34 ,  .1 ^2 )   \nonumber \\
V_0  & = &  .03 (not\ estimated) \nonumber
\end{eqnarray}
Since then, several other methods of have been used to calculate
BOLD parameters from FMRI timecourses. In \cite{Riera2004}, a maximum
likelihood method for innovation processes was used, as described by
\cite{Ozaki1994}. \cite{Ozaki1994} uses a similar construction to a 
Kalman filter, to break the time series into a series of innovations,
for which Maximum Likelihood was performed. While this is in some since the
"right" way to find the solution, it comes with several caveats. First, every
step in parameter space requires a recalculation of all the state variables. With
two or three parameters this is fine, more than that, and calculations could go on
indefinitely. Second, it still assumes the parameters and noise are Gaussian, and
will only be optimal in that case. Third, depending on the nonlinearities present
in the system, local minima may be extremely common. Later \cite{Hu2009} used an 
Unscented Kalman Filter over all the parameters and state variables to find the 
parameter set/variable time series. While this method has the drawback of not necessarily
being optimal, unfortunately there is no general optimal solution to non-linear non
Gaussian models. Hu et. al.'s technique also will run significantly faster than
ML based techniques, since it does not require recalculating the entire timeseries
for every step in parameters. Both \cite{Hu2009} and \cite{Friston2002} came to results 
very similar to the expected values stated in \cite{Buxton1998} and \cite{Friston2000}.
One potential problem with all these techniques are that the depend heavily on the
priors. The starting point of the parameters could have a huge impact on the
results, and while one can be hopeful that this isn't the reason for the agreement
between \cite{Friston2000} and later results, there is no way to know.

In \cite{Jonston2008}, a hybrid particle filter/gradient
descent algorithm was used to simultaneously derive the static (classicly called
parameters) and dynamic parameters (classically known as state variables).
Essentially a particle filter is used to calculate the state variables at some
time, and then the estimated distribution of the particles was used to find
the most likely set of parameters that would give that distribution of state variables.
\cite{Johnston2008} comes to a very different set of parameter estimates as compared
to the original \cite{Friston2000} guesses:
\begin{eqnarray}
epsilon &=& .069 \pm .014    \nonumber \\
tau_s & =& 4.98 \pm 1.07  \nonumber \\
tau_f & =& 8.31 \pm 1.51   \nonumber \\
tau_0 & =& 8.38 \pm 1.5    \nonumber \\
alpha & =& .189 \pm .004   \nonumber \\
E_0   & =& .635 \pm .072     \nonumber \\
V_0   & =& .0149 \pm .006     \nonumber 
\end{eqnarray}
Notably, the time constants are significantly longer. This could be a result of
some preprocessing to the stimulus timeseries performed in \cite{Friston2002} and
later works but not in \cite{Johnston2008}, or it could be that \cite{Johnston2008}
depended less on the priors.

@@

\subsection{Noise}
Thus, despite some discrepencies, the cause of the BOLD signal is
relatively well known.  but FMRI doesn't detect this happening in one neuron,
but rather as the 
aggregate over millions of cells. Though local neurons act
"together" (i.e. around the same time), the density of neurons, the
density of capillaries, and slight differences in activation across 
a particular voxel can all lead to signal attenuation or noise. 
A particularly insidious type of noise present in FMRI is a low frequncey
drift, first characterized by a Weiner process in \cite{Riera2004}. 
Though not present in all regions, it is prevalent enough to cause significant
inference problems \cite{Tanabe2002}. It is still not
clear where exactly this noise comes from, although it is possible it is 
the result of magnets heating up, or some distortion in magnetic
fields \cite{Smith2007}. It is clear that this drift signal is not soley
due to a physiological
effects, given its presence in cadavers and phantoms \cite{Smith1999}.
Additionally, since it is standard operating procedure to perform
coregistration between volumes, it is unlikely that movement was
the cause of the drift either (otherwise the effect would not be 
widely reported).

In order to characterize the noise in the FMRI we took a few 
sample time-series during resting state, which should theoretically be all noise and 
used Q-Q plots to analyze the type of noise present. \ref{fig:QQDC} shows the 
results of a Q-Q plots between the samples and a Gaussian of mean and variance based
on the samples. Because it the noise is often considered to have Weiner noise with
Gaussian steps, \ref{fig:QQDelta} shows a Q-Q plot using the steps, rather than 
absolute values.
\begin{figure}
\label{fig:QQDC}
\caption{Q-Q Plots of resting state data}
\end{figure}

\begin{figure}
\label{fig:QQDelta}
\caption{Q-Q Plots of resting state data, using the steps as the samples}
\end{figure}

Finally, to determine whether much of the autocorrelation was removed by taking
steps we plotted the autocorrelation of each sample location, before for the
actual value (\ref{fig:CorrDC}), and the steps between values (\ref{fig:CorrDelta}).
Ideally, if the steps are Gaussian, they should be independent and have a
flat autocorrelation; although in real life this is far too much to expect.

\begin{figure}
\label{fig:CorrDC}
\caption{Autocorrelation of the absolute value of various time-series}
\end{figure}

\begin{figure}
\label{fig:CorrDelta}
\caption{Autocorrelation of the steps of various time-series}
\end{figure}

@@

\subsection{Analysis of the BOLD Signal}
Several studies have endeavored to analyze the BOLD equations, which
we will review here. The most obvious analysis is to calculate the
steady state signal, when $u$ is held constant long enough for
all the transients to settle out.

\begin{eqnarray}
s_{ss} &=& 0 \nonumber \\
f_{ss} &=& \tau_f\epsilon u + 1\nonumber \\
v_{ss} &=& (\tau_f\epsilon u + 1)^\alpha\nonumber \\
q_{ss} &=& \frac{(\tau_f\epsilon u + 1)^\alpha}{E_0}(1-(1-E_0)^{1/(\tau_f\epsilon u + 1)})\nonumber \\
y_{ss} &=& V_0((k_1+k_2)(1-q_{ss}) - (k_2+k_3)(1-v_{ss}))
\label{steadystate}
\end{eqnarray}

@@

\section{Current Techniques}
\subsection{Basic Statistical Parametric Mapping}
Although not strictly the same thing as parameter calculation from 
FMRI, activation detection is very similar. In fact, estimation of 
parameters is somewhat a generalization of the idea of activation detection.
Thus, it is important to draw a comparison between the methods proposed
in this thesis with existing methods of activation detection.

The most basic method of analyzing FMRI data is through a standard T-test
between "resting state" and "active state" samples. This is done by 
taking the average and variance of the inactive period, and the 
period during which the stimulus was activate separately then treating 
them both as gaussian distributions.
If they are in fact Gaussian distributions, then a basic t-test will
give the probability that the samples came from the same distribution
(the null hypothesis). Of course, this test is fraught with problems; even if
the drift mentioned earlier has been removed, there is little reason
to believe that the noise is Gaussian, or even stable. Additionally, 
even if the noise were Gaussian, a t-test with a p-value of .05 over
50000 or more samples is on average going to generate $.05*50000$ false
positives. To compensate for this, bonferoni correction, also known as
multiple comparison tests are performed; essentially p-values are 
divided by the number of independent
tests being run. This, however, leads to extremely low p-values, so
low that it would be impossible for any biological system to satisfy. To
compensate, a Gaussian kernel is applied to the image, thus reducing
variance (and thus separating the active and inactive distributions)
as well as decreasing the effective number of voxels. Since t-tests are
now no longer being applied to n <I need to define n> independent voxels,
the factor by which the p-value must be divided by can be decreased.
<Do I need to mathematically define all this?> The derivation and application
of random field theory, and its use can be found in various papers \cite{univ_mult_rft}.

\subsection{General Linear Model}
The most used form of FMRI analysis is Statistical Parametric
Mapping, but is able to account for several different levels or types
of stimulus (see \cite{Hofmann1997}). By using hierarchical models
the output signal timeseries is considered the weighted
sum of the various input timeseries. Essentially every experimental factor
is considered as another level inputs (ex. different type of stimulus,
different run with the same patient, different patient). 
The equation for a general linear model is then
\begin{equation}
Y(t) = X(t)\beta + \epsilon(t)
\end{equation}
where $Y(t)$ is the smoothed or detrended timeseries of measurements,
$X(t)$ is a row vector of input, $\beta$ is a column vector of weights,
and $\epsilon$ is the error. Thus for every time, the measurement is
assumed to be a weighted sum of the columns of $X$ plus some error. The calculation
of $\beta$ is then performed using a maximum likelihood or gradient descent search 
to minimize the error.

<Image of GLM>

@@

<Image of Hemodynamic Response Function>

@@

\begin{figure}
\label{fig:EKFWorking}
\caption{Examples of update distributions, using an Kalman Filter, \cite{Thrun}}
\end{figure}

>>>>

\documentclass{report}
\usepackage[left=3cm,top=2cm,right=3cm]{geometry}

\usepackage{times}
\usepackage[siunitx]{circuitikz}
%\usetikzlibrary{trees}
%\usetikzlibrary{shapes,snakes}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{cancel}
\usepackage[colorlinks=true]{hyperref}

@@

\begin{document}
\chapter{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
of neural activity. Despite its rather limited temporal resolution, FMRI is still
the best way of measuring neural activity for the majority of the brain.
Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis straight forward.
Because of these benefits, FMRI continues to be crucial to the study of human 
cognition. Despite its prevalence, there have been relatively few developments
in the actual analysis of FMRI images. A steady stream of studies have built
on the original BOLD signal derivation first described in \cite{Ogawa}, 
from the Baloon model first proposed by \cite{Buxton1998}
all the way to full fully autonomous system of equations \cite{Riera2004}. And while
there have been numerous forks in the model, enough in fact to make an entire paper
studying the differences,\cite{Deneux2006}, it is widely known that all these
models have quantitatively less bias error than General Linear Model which is
typically employed today. Then again, depending on the model there may be between
seven \cite{Riera2004} and 50 \cite{Behzadi2005} parameters per voxel to
be optimized. Clearly there is a significant risk of error due to variance
with so many degrees of freedom, not to mention a significantly increased
computation cost. In this thesis I demonstrate the use of a 
particle filter as a means of addressing these problems.

@@

\section{FMRI}
Magnetic Resonance Imaging, MRI is a method of building 3D images
non-invasively, based on the difference between nuclear spin
relaxation times in various molecules. Initially the entity being
imaged is brought into a large magnetic field which aligns the spins
of molecules in the same direction; radio frequency (RF) signals may
then be used to excite nuclear spin away from the steady alignment. 
As the nuclei precess back to their original orientation, they resonate
at the same RF frequency of their original excitation. Conveniently, the
excitation of nuclear spins return their original state at different
rates, called the T1 relaxation time, depending on the properties of 
the material excited. Additionally, the
coherence of the spins also decay differently (and quite a bit faster
than T1) based on the properties of the region that has been excited.
This gives two primary methods of contrasting substances,
which is the basis of T1 and T2 weighted images. Additionally, there
dephasing occurs at two different rates, the T2 relaxation time,
which is impossible to recover from, and T2$^*$ relaxation, which is
much faster, but possible to recover from with an inversion pulse.
Oftentimes T1 relaxation times can be on the order of seconds if 
a significant excitation pulse is applied. 
In order to rapidly acquire entire brain images, as is done in Functional 
MRI, a single large excitation pulse is applied to the entire brain,
and the entire volume is acquired in a single T1 relaxation period. 
Because the entire k-space (spatial-frequency) volume is acquired 
from a single excitation, the signal to noise ration is very low
in this type  of imaging (Echo Planar Imaging).

@@

\section{BOLD Physiology}
\label{sec:BOLD Physiology}
It is well known that the two types of hemoglobin act as a contrast agents in 
EPI imaging
\cite{Buxton1998}, \cite{WEISSKOFF1994}, \cite{Ogawa}, however the connection
between Deoxyhemoglobin/Oxygenated Hemoglobin and neural activity is non-trivial. 
Intuitively, increased 
metabolism will increase Deoxyhemoglobin, however blood vessels are quick
to compensate by increasing local blood flow. Increased inflow will of course
preceed increased outflow, and increased inflow is accomplished by loosening 
capilary beds. Both of these factors drive increased storage capacity.
Since the local MR signal depends on the ratio of Deoxyhemoglobin to Oxygenated
Hemoglobin, increased volume of blood can certainly effect this ratio if 
metabolism doesn't exactly match the increased inflow of oxygenated blood.
This was the impetus
for the ground breaking balloon model (\cite{Buxton1998}) and windkessel
model (\cite{Mandeville1999}). These models derive from first principals
the increased deoxyhemoglobin ratio and volume of capillaries based on a given flow.
These were the first two attempts to quantitatively account for the shape of the 
BOLD signal as a consequence of the lag between the cerebral blood volume (CBV) 
and the cerebral blood flow (CBF). In fact \cite{Buxton1998} went to far as
to show that a simple, well chosen blood flow waveform coupled with a square 
wave cerebral metabolic rate of oxygen (CMRO2) curve, in the context of a balloon 
model, could fully account for the BOLD signal.

Although \cite{Buxton1998} showed that a well chosen flow waveform could 
explain much of the BOLD signal, there was still a matter of proposing a
realistic waveform for the CBF and for the CMRO2. \cite{Friston2000} gave
a reasonable and simple
expressoin for CBF input,$f$, based on a flow inducing signal, $s$, 
\begin{eqnarray}
\dot{s} &=& \epsilon u(t) - \frac{s}{\tau_s} - \frac{f - 1}{\tau_f} \\
\dot{f} &=& s
\end{eqnarray}
where $\epsilon$ is a neuronal efficiency term, $u(t)$ is a stimulus, and $\tau_f$, $\tau_s$
are both time constants. In \cite{Buxton2004} the final piece of the simple balloon
model was put into place, by describing the CMRO2 as a constant multiple of
the CBF (the inflow of blood). This completed the basic balloon model, and 
was well summarized in \cite{Riera2004}. 
\begin{eqnarray}
\dot{v} &=& \frac{1}{\tau_0}(f - v^\alpha)\\
\dot{q} &=& \frac{1}{\tau_0}(\frac{f(1-(1-E_0)^f)}{E_0} - \frac{q}{v^{1-1/\alpha}})
\end{eqnarray}
where $v$ is normalized cerebral blood volume (CBV), and $q$ is the normalized
local deoxyhemoglobin/oxygenated hemoglobin ratio, $E_0$ is the resting metabolic
rate and $\alpha$ is Grubb's parameter controling the balloon model. 
\cite{Obata2004} refined the readout equation of the BOLD signal based on the
deoxyhemoglobin content (q) and local blood volume (v), resulting in the
final BOLD equation:
\begin{eqnarray}
y   &=& V_0((k_1 + k_2)(1-q) - (k_2 + k_3)(1-v))\\
k_1 &=& 4.3 \times \nu_0 \times E_0 \times TE = 2.8\\
K_2 &=& \epsilon_0 \times r_0 \times E_0 \times TE = .57\\
k_3 &=& \epsilon_0 - 1 = .43
\end{eqnarray}
Where $\nu_0 = 40.3 s^{-1}$  is the frequency offset in Hz for fully
deoxygenated blood (at 1.5T), $r_0 = 25 s^{-1}$  is the slope relating
change in relaxation rate with change in blood oxygenation, and
$\epsilon_0 = 1.43$ is the 
ratio of signal MR from intravascular to extravascular at rest. Although,
obviously these constants change with experiment ($TE$, $\nu_0$, $r_0$),
patient, and brain 
region ($E_0$, $r_0$), often the estimated values taken from \cite{Obata2004} are used
as constants ($k_1 + k_2 = 3.4$, and $k_2+k_3 = 1$) in 1.5 Tesla studies..
While this model is in a sense complete, it is far from perfect. The major
problem often brought up with this version of the BOLD model is that it
does not represent the so called "post-stimulus undershoot" well.
The post-stimulus undershoot is the name for a prolonged sub-normal
BOLD response for a period of 10 to 60 seconds after stimulus has
ceased (\cite{Chen2009}, \cite{Mandeville1999a}).

@@

Even more advanced versions of the Balloon model exist. In fact \cite{Buxton2004}
introduced several additional state variables, including the CMRO2, the O2 extraction
fraction, which is closely related to CMR02, and the neural response, which 
causes the stimulus to decay toward some steady state value. The 
neural response is intended to emulate neural habituation, wherein neurons
become less sensitive to a prolonged stimulus. While these advanced may be more
capable of capturing a more exact version of the BOLD signal, the difference
between the model will often times be below the noise floor. In essence this 
is a classic bias-variance
dilemma: at some point increased model flexibility, and thus variance, is not
worth the decrease in model bias. For now it remains to be seen where this
line may be drawn in the BOLD signal, although \cite{Deneux2006} does not
show a significant improvement from the additional parameters added in 
\cite{Buxton2004}.

\section{Previous Studies of Parameters}
There have been quite a few efforts to quantify the parameters of the
various BOLD models.  Although \cite{Buxton1998} and \cite{Friston2000}
both proposed physiologically reasonable values for the model parameters, 
\cite{Friston2002} was the first paper to calculate the parameters based 
on actual FMRI data. In that paper, Friston et. al. used a variation of
Expectation Maximization to find a normal distribution for the parameters:
\begin{eqnarray}
\epsilon &=& N(.54 , .1 ^2 )  \nonumber \\
\tau_s & =&  N(1.54, .25^2)   \nonumber \\
\tau_f & =&  N(2.46, .25^2)   \nonumber \\
\tau_0 & =&  N(.98 , .25^2 )   \nonumber \\
\alpha & =&  N(.33 , .45^2 )   \nonumber \\
E_0   & =&  N(.34 ,  .1 ^2 )   \nonumber \\
V_0  & = &  .03 (not\ estimated) \nonumber
\end{eqnarray}
Since then, several other methods of have been used to calculate
BOLD parameters from FMRI timecourses. In \cite{Riera2004}, a maximum
likelihood method for innovation processes was used, as described by
\cite{Ozaki1994}. \cite{Ozaki1994} uses a similar construction to a 
Kalman filter, to break the time series into a series of innovations,
for which Maximum Likelihood was performed. While this is in some since the
"right" way to find the solution, it comes with several caveats. First, every
step in parameter space requires a recalculation of all the state variables. With
two or three parameters this is fine, more than that, and calculations could go on
indefinitely. Second, it still assumes the parameters and noise are Gaussian, and
will only be optimal in that case. Third, depending on the nonlinearities present
in the system, local minima may be extremely common. Later \cite{Hu2009} used an 
Unscented Kalman Filter over all the parameters and state variables to find the 
parameter set/variable time series. While this method has the drawback of not necessarily
being optimal, unfortunately there is no general optimal solution to non-linear non
Gaussian models. Hu et. al.'s technique also will run significantly faster than
ML based techniques, since it does not require recalculating the entire timeseries
for every step in parameters. Both \cite{Hu2009} and \cite{Friston2002} came to results 
very similar to the expected values stated in \cite{Buxton1998} and \cite{Friston2000}.
One potential problem with all these techniques are that the depend heavily on the
priors. The starting point of the parameters could have a huge impact on the
results, and while one can be hopeful that this isn't the reason for the agreement
between \cite{Friston2000} and later results, there is no way to know.

In \cite{Johnston2008}, a hybrid particle filter/gradient
descent algorithm was used to simultaneously derive the static (classicly called
parameters) and dynamic parameters (classically known as state variables).
Essentially a particle filter is used to calculate the state variables at some
time, and then the estimated distribution of the particles was used to find
the most likely set of parameters that would give that distribution of state variables.
\cite{Johnston2008} comes to a very different set of parameter estimates as compared
to the original \cite{Friston2000} guesses:
\begin{eqnarray}
epsilon &=& .069 \pm .014    \nonumber \\
tau_s & =& 4.98 \pm 1.07  \nonumber \\
tau_f & =& 8.31 \pm 1.51   \nonumber \\
tau_0 & =& 8.38 \pm 1.5    \nonumber \\
alpha & =& .189 \pm .004   \nonumber \\
E_0   & =& .635 \pm .072     \nonumber \\
V_0   & =& .0149 \pm .006     \nonumber 
\end{eqnarray}
Notably, the time constants are significantly longer. This could be a result of
some preprocessing to the stimulus timeseries performed in \cite{Friston2002} and
later works but not in \cite{Johnston2008}, or it could be that \cite{Johnston2008}
depended less on the priors.

@@

\section{Noise}
Thus, despite some discrepencies, the cause of the BOLD signal is
relatively well known.  but FMRI doesn't detect this happening in one neuron,
but rather as the 
aggregate over millions of cells. Though local neurons act
"together" (i.e. around the same time), the density of neurons, the
density of capillaries, and slight differences in activation across 
a particular voxel can all lead to signal attenuation or noise. 
A particularly insidious type of noise present in FMRI is a low frequncey
drift, first characterized by a Weiner process in \cite{Riera2004}. 
Though not present in all regions, it is prevalent enough to cause significant
inference problems \cite{Tanabe2002}. It is still not
clear where exactly this noise comes from, although it is possible it is 
the result of magnets heating up, or some distortion in magnetic
fields \cite{Smith2007}. It is clear that this drift signal is not soley
due to a physiological
effects, given its presence in cadavers and phantoms \cite{Smith1999}.
Additionally, since it is standard operating procedure to perform
co-registration between volumes, it is unlikely that movement was
the cause of the drift either (otherwise the effect would not be 
widely reported).

In order to characterize the noise in the FMRI we took a few 
sample time-series during resting state, which should theoretically be all noise and 
used Q-Q plots to analyze the type of noise present. Additionally, the autocorrelation
was also calculated since most methods (including the one used in this paper)
assume the noise is independent from time and
identically distributed.  If these assumptions fail, then error rate will
be higher than expected. \autoref{fig:QQDC} shows the 
results with a regression line fit to the points. A set of points that are exactly 
Gaussian should all fall on the same line.  Because it the noise is often considered 
to have Weiner noise with Gaussian steps, as first described in \cite{Riera2003}, we
also performed the same tests again, but on the steps, rather than the 
direct measurements. \autoref{fig:QQDelta} shows the analysis of the same
set of time-series as \autoref{fig:QQDC}, after being converted to steps.
Finally, as we will discuss in \autoref{sec:Methods Preprocessing}, some
amount of drift can be removed by subtracting off a spline from the timeseries.

\begin{figure}
\caption{Q-Q Plots of resting state data}
\label{fig:QQDC}
\end{figure}

\begin{figure}
\caption{Q-Q Plots of resting state data, using the steps as the samples}
\label{fig:QQDelta}
\end{figure}

@@

\section{Analysis of the BOLD Signal}
Several studies have endeavored to analyze the BOLD equations, which
we will review here. The most obvious analysis is to calculate the
steady state signal, when $u$ is held constant long enough for
all the transients to settle out.

\begin{eqnarray}
s_{ss} &=& 0 \nonumber \\
f_{ss} &=& \tau_f\epsilon u + 1\nonumber \\
v_{ss} &=& (\tau_f\epsilon u + 1)^\alpha\nonumber \\
q_{ss} &=& \frac{(\tau_f\epsilon u + 1)^\alpha}{E_0}(1-(1-E_0)^{1/(\tau_f\epsilon u + 1)})\nonumber \\
y_{ss} &=& V_0((k_1+k_2)(1-q_{ss}) - (k_2+k_3)(1-v_{ss}))
\label{eq:steadystate}
\end{eqnarray}

@@

\chapter{Current Techniques}
\section{Basic Statistical Parametric Mapping}
Although not strictly the same thing as parameter calculation from 
FMRI, activation detection is very similar. In fact, estimation of 
parameters is somewhat a generalization of the idea of activation detection.
Thus, it is important to draw a comparison between the methods proposed
in this thesis with existing methods of activation detection.

The most basic method of analyzing FMRI data is through a standard T-test
between "resting state" and "active state" samples. This is done by 
taking the average and variance of the inactive period, and the 
period during which the stimulus was activate separately then treating 
them both as gaussian distributions.
If they are in fact Gaussian distributions, then a basic t-test will
give the probability that the samples came from the same distribution
(the null hypothesis). Of course, this test is fraught with problems; even if
the drift mentioned earlier has been removed, there is little reason
to believe that the noise is Gaussian, or even stable. Additionally, 
even if the noise were Gaussian, a t-test with a p-value of .05 over
50000 or more samples is on average going to generate $.05*50000$ false
positives. To compensate for this, bonferoni correction, also known as
multiple comparison tests are performed; essentially p-values are 
divided by the number of independent
tests being run. This, however, leads to extremely low p-values, so
low that it would be impossible for any biological system to satisfy. To
compensate, a Gaussian kernel is applied to the image, thus reducing
variance (and thus separating the active and inactive distributions)
as well as decreasing the effective number of voxels. Since t-tests are
now no longer being applied to n <I need to define n> independent voxels,
the factor by which the p-value must be divided by can be decreased.
<Do I need to mathematically define all this?> The derivation and application
of random field theory, and its use can be found in various papers \cite{Worsley2004}.

\section{General Linear Model}
The most used form of FMRI analysis is Statistical Parametric
Mapping, but is able to account for several different levels or types
of stimulus (see \cite{Hofmann1997}). By using hierarchical models
the output signal timeseries is considered the weighted
sum of the various input timeseries. Essentially every experimental factor
is considered as another level inputs (ex. different type of stimulus,
different run with the same patient, different patient). 
The equation for a general linear model is then
\begin{equation}
Y(t) = X(t)\beta + \epsilon(t)
\end{equation}
where $Y(t)$ is the smoothed or detrended timeseries of measurements,
$X(t)$ is a row vector of input, $\beta$ is a column vector of weights,
and $\epsilon$ is the error. Thus for every time, the measurement is
assumed to be a weighted sum of the columns of $X$ plus some error. The calculation
of $\beta$ is then performed using a maximum likelihood or gradient descent search 
to minimize the error.

[Image of GLM]

@@

[Image of Hemodynamic Response Function]

@@

\begin{figure}
\caption{Examples of update distributions, using an Kalman Filter, \cite{Thrun2005}}
\label{fig:EKFWorking}
\end{figure}

The difficulty
of using a Kalman Filter, however, is that it assumes a multivariate 
Gaussian for the state variables, $X(t-1)$. The problem with this is that 
when a system is significantly nonlinear, the state at 
$X(t)$ will almost certainly be non-Gaussian, and thus estimating
$X(t+1)$ with $X(t)$ as a multivariate Gaussian will perpetually introduce
error in the distribution. Furthermore, it is not really known what 
sort of underlying distributions may exist in such a mixed biological,
mechanical, chemical system such as the brain. Assuming the parameters
all to be Gaussian may in fact be a gross error. On the other hand, for
small variances and short time steps the gaussiann distribution is a good 
fit, and so in some limited cases the Unscented Kalman Filter could work
well. These are 
non-trivial issues given that the assumption of Gaussianity is what allows
the UKF to estimate the posterior using only the first and second moments;
two parameters that don't uniquely describe most distributions.
